Self-reported LLM usage and outcomes on a data science project: Evidence from two undergraduate data science courses

### 2025-04-05 paper/paper.qmd
Add SSTA304

---
title: "Self-reported LLM usage and outcomes on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
  - Nathalie Moon
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is was approved by the University of Toronto's Research Ethics Board as Protocol 47725. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which self-reported LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest is similar to the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

library(arrow)
library(here)
library(knitr)
library(tidyverse)
library(scales)
library(modelsummary)
library(bayesplot)

# Read in data
all_data <-
  read_parquet(here("data/analysis_data.parquet"))
```

@@

Trustworthy data science is the practice of conducting data analysis in a transparent, ethical, and reliable manner. These principles are upheld in various ways, through ongoing education, adherence to professional norms and implementation of reproducible workflows. To stay current, data scientists must continually update their knowledge of best practices for transparency and reproducibility, foster a culture which values different perspectives and accountability, and critically consider the broader impact of data-driven decisions. Recent advances, such as the wide release of user-friendly Large Language Models (LLMs), particularly OpenAI’s ChatGPT, have transformed the data science toolkit. These powerful tools for natural language processing and generation influence various aspects of data analysis and automation, further emphasizing the need for trustworthy practices.

Like all new tools LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. LLMs have quickly been put to use in both industry and academia. By now many people, especially educators and students, have some experience with LLMs in both personal and professional contexts.

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [@fulgencio2024developing; @afzal2024tailoring]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Eke2023], the development of critical thinking skills [@thiga2024generative], and what effective learning looks like [@BaidooAnu_2023].

The tasks involved in a trustworthy data science workflow can be generally broken down into a number of key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. One is programming, which is done when cleaning, analyzing, and visualizing data often using programming languages like R or Python. Another is writing, which is primarily done when communicating results. The potential for LLMs to positively affect students' academic performance in data science follows from their already-demonstrated capability in those competencies in adjacent professional fields.

In terms of computer programming, @Peng_2023 found a positive impact of GitHub Copilot (an LLM-powered programming assistant) on productivity. Specifically, in an experiment involving 95 freelance programmers, they found that a treatment group of programmers with access to GitHub Copilot completed a standardized programming task 56% faster than the control group. Programmers with less experience saw the greatest improvements in productivity.

@DellAcqua_2023, focusing on management consulting tasks, provide evidence that LLMs can improve writing productivity. In a field experiment involving consultants from Boston Consulting Group they found that the use of OpenAI's GPT-4 led to a 25% increase in delivery speed of business tasks, most of which involved some writing, as well as a 40% increase in human-rated performance on those tasks. Similar to computer programming, these productivity increases were most pronounced for those with below average performance, with their output increasing by 43%.

On the other hand, @Valenzuela_2024 argue that LLMs lead to a loss of serendipity which may lead to less original work, and potentially de-skilling primarily with respect to programming ability, among other consequences. These outcomes could negatively affect students' effective learning of data science.

@Ellis_Slade_2023 take a more optimistic perspective and argue that LLMs are just another technology that will impact statistics and data science education. Similarly, @Tu_2024 acknowledge that LLMs can streamline many parts of a data science workflow. With that in mind, they suggest that data-scientists-in-training should shift their self-perspectives from primarily being an "analyst" to primarily being a "product manager" responsible for strategic oversight of the analysis carried out by LLMs.

At the high school level @Lazar_2023 conducted an informal survey of secondary school teachers and students on their opinions of ChatGPT, and found that LLMs could help creativity, provide academic support when teachers were unavailable, and model certain types of writing well. But teachers were concerned about the potential of LLMs to limit students' learning in certain ways through over-reliance. Beyond academic integrity concerns, @Lazar_2023 found that teachers had similar concerns to @Valenzuela_2024 about de-skilling and an overall loss of agency in writing and critical thinking.

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward, and usage of, AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. In particular, only 11% "strongly agreed" that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. Many respondents felt that using it to write whole papers was inappropriate, but using it for basic tasks like general assistance, writing feedback and basic data visualization appropriate.

In this paper we are interested in better understanding LLMs as a tool for producing trustworthy data science. We study how they were used by students in an upper-year undergraduate data science course and whether students who used LLMs tended to have higher scores than those who did not.

To understand the current state of LLMs as a tool for trustworthy data science, this paper focuses on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between the mark a student got on a final paper and self-reported measures of student LLM usage, as well as student attitudes toward LLMs in general. This is based on students' final papers and a survey, conducted in a third-year undergraduate data science course at the University of Toronto. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, current practice with regard to LLM integration in data science can be better understood, leading to better recommendations for their future development.

@@

All data are from the cohort of students taking STA302 "Methods of Data Analysis I" in the Winter 2024 semester at the University of Toronto. This course had 275 students initially enrolled which, reflecting a normal rate of attrition for undergraduate statistics courses at the University of Toronto, reduced to 154 students by the end of the semester. Assessment was heavily based on three papers submitted over the course of the 12 week semester.

The student marks that we analyze are based only on the final paper, which is done individually. By this stage, uninterested students have typically dropped the course, and students are familiar with course expectations. A typical paper submission is 10-20 pages, and requires students to conduct original research to answer a research question of interest to them. It reflects the skills typically used by a professional data scientist. Students are expected to develop a research question of interest to them, identify or collect data to answer the question, conduct statistical analysis, and write a short paper. Examples of final papers (shared with consent) include: @hannahyu; @emilysu; and @bennyrochwerg.

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due, students have received feedback on all their previous work in the class (including their past papers) and there is an optional two-day period of peer review.

@@

Throughout the semester students were encouraged to use LLMs. Formal instruction was provided twice during the semester. The first was a masterclass taught by a computer science faculty member on the ethics of using LLMs (see @Horton2024 for details). The second was a masterclass taught by a TA on writing with LLMs.

Data were collected from students through an optional end-of-course survey. [Appendix -@sec-survey-details] details the questions asked in the survey. Whether or not they consented to their data being used, all respondents received a 1% increase in their final course grade for their participation. Consenting responses were then matched to their final paper mark, as well as the GitHub repository for their final paper. The responses were then anonymized by removing any personal references to the students themselves including, names, emails, student numbers, GitHub links, and summarizing free-text responses.

@@

There were 146 responses to the survey. Of these, 119 respondents provided authorization for their data to be collected and used. Four of those respondents submitted the survey twice, and after removing their second response, 115 responses remained. Of those, 15 respondents did not include a statement on LLM usage in the README of the GitHub repository of their final paper, leaving 100 responses. Finally 7 of those respondents did not provide a usable GPA response.

All but one respondent completed the survey within 45 minutes (@fig-response-time-and-gpa-1). That one respondent took more than 4,000 minutes to complete the survey, suggesting they took a break while filling it out. After that respondent is removed from the analysis the average time to complete the survey was 11 minutes, and the standard deviation was 8 minutes. This leaves 92 respondents that were of use and were merged based on student name.

```{r fig-response-time-and-gpa}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of respondents' survey response times and self-reported GPA
#| fig-subcap: ["Distribution of survey response times","\"What is your GPA?\""]
#| layout-ncol: 1

all_data |>
  # summarize(mean = mean(mins_complete/60),
  # standard_dev = sd(mins_complete/60))
  ggplot(aes(x = mins_complete/60)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Time to complete survey (minutes)", y = "Frequency") +
  theme_minimal()

all_data |>
  # summarize(mean = mean(what_is_your_gpa, na.rm = TRUE),
  #           standard_dev = sd(what_is_your_gpa, na.rm = TRUE))
  ggplot(aes(x = what_is_your_gpa)) +
  geom_histogram(binwidth = 0.1) +
  labs(x = "Self-reported GPA", y = "Frequency") +
  theme_minimal()
```

There is a wide distribution of self-reported GPAs (@fig-response-time-and-gpa-2). The majority of responses cluster around a B (3.0 out of 4.0), and the average is 3.06 with a standard deviation of 0.55. One factor that may affect the range is that the course is required for programs in the Statistics, Mathematics and Computer Science Departments. While there was no reason for the respondents to not report the truth, self-reported GPAs also introduce the possibility of reporting bias. For instance, respondents may have provided their cumulative GPA, their most recent term's GPA, or could have misreported it entirely.

Students from a range of years took the course, however the majority of respondents were in their 3rd or 4th year of study (@tbl-year-dist). The course's prerequisite two-course sequence is typically completed by students in their second year, would make it difficult to take this course earlier than their 3rd year.

```{r tbl-year-dist}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: \"What year are you?\"

all_data |>
  count(what_year_are_you) |>
  pivot_wider(
    names_from = what_year_are_you,
    values_from = n,
    values_fill = list(n = 0)
  ) |>
  kable()
```

Respondents had a varied self-perception of their coding and writing abilities (@fig-selfperception). Most respondents believe it is important to be good at writing, but many are either indifferent or do not like to write (@fig-selfperception-1). Respondents also do not find writing to be particularly easy, which could be associated with the reported relative antipathy toward writing. Most respondents were at least somewhat confident in their own writing abilities, but a substantial number felt otherwise. Although few respondents felt that they were confident in their writing ability, more felt that they were able to catch their mistakes, which could indicate a disconnect between how respondents perceive their work and how the work was evaluated.

```{r fig-selfperception}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Self-perception of coding and writing abilities
#| fig-subcap: ["\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\"", "\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\""]

@@

Respondent self-perceptions regarding coding proficiency and importance varied (@fig-selfperception-2). There is strong consensus on the perceived importance of coding skills. However, respondents' self-assessed coding abilities and enjoyment are more heterogeneous, with a substantial proportion reporting moderate rather than high levels of ease and enjoyment in coding tasks.

Overall there was a moderate level of confidence among respondents in their overall coding ability, willingness to share code, and capacity to identify errors (@fig-selfperception-2). Notably, respondents express slightly higher confidence in detecting their own coding mistakes compared to general coding ability or code sharing. These patterns suggest that while respondents have developed some coding self-efficacy, there is still considerable potential for enhancing their perceived competence and comfort across various coding-related activities.

The majority of respondents were at least "somewhat familiar" with generative AI such as ChatGPT (@tbl-ai-familiarity-usage-1). Though respondents' self-perceptions around writing and coding were varied, there was strong consensus that the use of generative AI tools is appropriate within an academic setting (@tbl-ai-familiarity-usage-2). Most respondents who selected "It depends" generally found artificial intelligence tools to be appropriate, though with certain guidelines and rules governing their use.

```{r tbl-ai-familiarity-usage}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Familiarity with, and appropriateness of, generative AI
#| tbl-subcap: ["'How familiar are you with using generative AI tools such as OpenAI's ChatGPT or equivalents?'","'To what extent do you think using generative AI tools such as ChatGPT by OpenAI (or equivalents) is ethical and appropriate for schoolwork?'"]

all_data |>
  count(ai_familiarity) |>
  kable(col.names = c("Extent of AI familiarity", "Students"))

all_data |>
  count(ai_schoolwork_appropriate) |>
  kable(col.names = c("Ethical & Appropriate for school?", "Students"))
```

To understand the role of LLMs in learning, respondents identified their usage in a more granular way by selecting various pre-defined use cases in the survey (@fig-AI-use-and-usefulness-1). Technical questions and explaining concepts were the two top use cases among respondents. More than half of respondents also used LLMs for quick questions, general knowledge, writing paper code, and checking solutions. Just less than half reported that they used it to write paper content, which could suggest that respondents do not feel confident using it to improve writing.

```{r fig-AI-use-and-usefulness}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: "Use and usefulness of generative AI"
#| fig-subcap: ["\"If you have used generative AI tools such as OpenAI's ChatGPT or equivalents, in what ways have you used it (select all that apply)?\"", "\"How helpful did you find generative AI tools such as ChatGPT by OpenAI (or equivalents) for each component of STA302?\""]

# Use
all_data |>
  select(how_have_you_used) |>
  separate_rows(how_have_you_used, sep = ";") |>
  mutate(
    how_have_you_used = case_when(
      how_have_you_used == "check grammar mistake" ~ "Other",
      how_have_you_used == "debugging code" ~ "Other",
      how_have_you_used == "mental health support" ~ "Other",
      how_have_you_used == "Making practice questions for me with increasing difficulty" ~ "Other",
      how_have_you_used == "Generating test questions" ~ "Other",
      TRUE ~ how_have_you_used
    )
  ) |> 
  count(how_have_you_used) |>
  filter(how_have_you_used != "") |>
  arrange(desc(n)) |>
  ggplot(aes(x = n, y = reorder(how_have_you_used, n))) +
  geom_bar(stat = "identity") +
  labs(x = "Number of respondents", y = "Type of usage") +
  theme_minimal()

>>>>

---
title: "Self-reported LLM usage and outcomes on a data science project: Evidence from two undergraduate data science courses"
author: 
  - Rohan Alexander
  - Luca Carnegie
  - Nathalie Moon
thanks: "Code and data are available at: REDACTED. We thank attendees at JSM 2024 for helpful suggestions. This research was approved by the University of REDACTED's Research Ethics Board (Protocol 47725). Comments: REDACTED."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which self-reported LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the task of interest is similar to the work done by professional data scientists. Comparing self-reported usage, there was a considerable increase between the class in January-April 2024 where 41 per cent of students self-reported extensive LLM usage and the class in September-December 2024 where 69 per cent reported extensive LLM usage. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

library(arrow)
library(here)
library(knitr)
library(tidyverse)
library(scales)
library(modelsummary)
library(bayesplot)
library(tidybayes)

# Read in data
all_data <-
  read_parquet(here("data/analysis_data.parquet"))

sta304_data <-
  read_parquet(here("data/sta304_analysis_data_classified_cleaned.parquet"))

# Read in models
fit1 <-
  readRDS(file = here("models/fit_one_inflated.rds"))
fit2 <-
  readRDS(file = here("models/fit_one_inflated_gpa.rds"))
fit3 <-
  readRDS(file = here("models/fit_one_inflated_sta304.rds"))
fit4 <-
  readRDS(file = here("models/fit_one_inflated_gpa_sta304.rds"))
fit5 <-
  readRDS(file = here("models/fit_one_inflated_gpa_english_sta304.rds"))
fit6 <-
  readRDS(file = here("models/fit_one_inflated_gpa_both.rds"))
```

@@

Trustworthy data science is the practice of conducting data analysis in a transparent, ethical, and reliable manner. These principles are upheld in various ways, through ongoing education, adherence to professional norms and implementation of reproducible workflows. To stay current, data scientists must continually update their knowledge of best practices for transparency and reproducibility, foster a culture which values different perspectives and accountability, and critically consider the broader impact of data-driven decisions. Recent advances, such as the wide release of user-friendly Large Language Models (LLMs), particularly OpenAI's ChatGPT, have transformed the data science toolkit. These powerful tools for natural language processing and generation influence various aspects of data analysis and automation, further emphasizing the need for trustworthy practices.

Like all new tools LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. LLMs have quickly been used in both industry and academia. By now many people, especially educators and students, have some experience with LLMs in both personal and professional contexts.

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [@fulgencio2024developing; @afzal2024tailoring]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Eke2023], the development of critical thinking skills [@thiga2024generative], and what effective learning looks like [@BaidooAnu_2023; @Bastani2024].

The tasks involved in a trustworthy data science workflow can be generally broken down into a few key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. One is programming, which is done when cleaning, analyzing, and visualizing data often using programming languages like R or Python. Another is writing, which is primarily done when communicating results. The potential for LLMs to positively affect students' academic performance in data science follows from their already-demonstrated capability in those competencies in adjacent professional fields.

In terms of computer programming, @Peng_2023 found a positive impact of GitHub Copilot (an LLM-powered programming assistant) on productivity. Specifically, in an experiment involving 95 freelance programmers, they found that a treatment group of programmers with access to GitHub Copilot completed a standardized programming task 56 per cent faster than the control group. Programmers with less experience saw the greatest improvements in productivity. @DellAcqua_2023, focusing on management consulting tasks, provide evidence that LLMs can improve writing productivity. In a field experiment involving consultants from Boston Consulting Group they found that the use of OpenAI's GPT-4 led to a 25 per cent increase in delivery speed of business tasks, most of which involved some writing, as well as a 40 per cent increase in human-rated performance on those tasks. Like computer programming, these productivity increases were most pronounced for those with below average performance, with their output increasing by 43 per cent.

On the other hand, @Valenzuela_2024 argue that LLMs lead to a loss of serendipity which may lead to less original work, and potentially de-skilling primarily with respect to programming ability, among other consequences. These outcomes could negatively affect students' effective learning of data science. @Ellis_Slade_2023 take a more optimistic perspective and argue that LLMs are just another technology that will impact statistics and data science education. Similarly, @Tu_2024 acknowledge that LLMs can streamline many parts of a data science workflow. With that in mind, they suggest that data-scientists-in-training should shift their self-perspectives from primarily being an "analyst" to primarily being a "product manager" responsible for strategic oversight of the analysis carried out by LLMs. @lehmann2025aimeetsclassroomlarge found the effect of LLMs on learning was nuanced and depended on the way the LLM was used and the level of prior knowledge. At the high school level @Lazar_2023 conducted an informal survey of secondary school teachers and students on their opinions of ChatGPT, and found that LLMs could help creativity, provide academic support when teachers were unavailable, and model certain types of writing well. But teachers were concerned about the potential of LLMs to limit students' learning in certain ways through over-reliance. Beyond academic integrity concerns, @Lazar_2023 found that teachers had similar concerns to @Valenzuela_2024 about de-skilling and an overall loss of agency in writing and critical thinking.

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward, and usage of, AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. Only 11 per cent "strongly agreed" that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. Many respondents felt that using it to write whole papers was inappropriate, but using it for basic tasks like general assistance, writing feedback and basic data visualization appropriate.

In this paper we are interested in better understanding LLMs as a tool for producing trustworthy data science. We study how they were used by students in two upper-year undergraduate data science courses -- one in Winter 2024 (January to April) and another in Fall 2024 (September to December) -- and whether students who used LLMs tended to have higher scores than those who did not. We focus on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between the mark a student got on a final paper and self-reported measures of student LLM usage, as well as student attitudes toward LLMs in general. This is based on students' final papers and a survey, conducted in third-year undergraduate data science courses at the University of REDACTED. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, current practice with regard to LLM integration in data science can be better understood, leading to better recommendations for their future development.

We find some mild association between the mark that a respondent received and their self-reported LLM usage. And this is especially the case for respondents that were not native English speakers. More concretely, there was a considerable increase in the number of respondents who self-reported extensive LLM usage and also those who considered LLM usage to be appropriate and ethical for school, when we compared the results in the class at the start of 2024 with the class at the end of 2024. Taken together it is clear that considerable additional work is needed understanding the appropriate use of generative AI in statistics and data science pedagogy.

@@

All data are from students taking STA302 "Methods of Data Analysis I" in the Winter 2024 semester (January to April 2024), or students taking STA304 "Surveys, Sampling, and Observational Data" in the Fall 2024 semester (September to December 2024), at the University of REDACTED. STA302 had 275 students initially enrolled which reduced to 154 students by the end of the semester. Similarly, STA304 had 275 students initially, reduced to 196 students by the end of the semester. These reductions reflect a normal rate of attrition for undergraduate statistics courses at the University of REDACTED. Assessment in both courses was heavily based on three papers submitted over the course of the 12 week semester.

The student marks that we analyze are based only on the final paper, which was done individually. By this stage, uninterested students have typically dropped the course, and students are familiar with course expectations. A typical paper submission was 10-20 pages, and required students to conduct original research to answer a research question of interest to them. It reflects the skills typically used by a professional data scientist. Students are expected to develop a research question of interest to them, identify or collect data to answer the question, conduct statistical analysis, and write a short paper. Examples of final papers (shared with consent) include: REDACTED; REDACTED; and REDACTED.

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due, students have received feedback on all their previous work in the class (including their past papers) and there is an optional period of peer review.

@@

Throughout the semester students were encouraged to use LLMs. In STA302 formal instruction in LLM usage was provided twice during the semester. The first was a masterclass taught by a computer science faculty member on the ethics of using LLMs (see @Horton2024 for details). The second was a masterclass taught by a TA on writing with LLMs. In STA304 formal instruction was provided by instructor-demonstration of efficient LLM usage for coding and writing.

Data were collected from students through an optional end-of-course survey. [Appendix -@sec-survey-details] details the questions asked in the survey for STA302. The STA304 survey was almost identical, with the one difference discussed later. Whether or not they consented to their data being used, all respondents received a 1 per cent increase in their final course grade for filling out the survey. Consenting responses were then matched to their final paper mark, as well as the GitHub repository for their final paper. The responses were then anonymized by removing any personal references to the students themselves including, names, emails, student numbers, GitHub links, and summarizing free-text responses.

@@

In STA302, there were 146 responses to the survey. Of these, 119 respondents provided authorization for their data to be collected and used. Four of those respondents submitted the survey twice, and after removing their second response, 115 responses remained. Of those, 15 respondents did not include a statement on LLM usage in the README of the GitHub repository of their final paper, leaving 100 responses. Finally, seven of those respondents did not provide a usable GPA response.

In STA304, there were 183 responses to the survey. Of these, 153 respondents provided authorization for their data to be collected and used. Three of those respondents submitted the survey twice and their second response was removed. One student submitted their final paper too late to be of use, and so their survey was removed. Nine respondents did not include a statement on LLM usage in the README of the GitHub repository of their final paper. One respondent did not provide a usable GPA response. Finally, one respondent failed an attention check question halfway through the survey "Please select the option blue".

In STA302, all but one respondent completed the survey within 60 minutes. After that respondent was removed from the analysis the average time to complete the survey was 11 minutes, and the standard deviation was 8 minutes (@fig-response-time-and-gpa-1). This left 92 respondents from STA302 that were of use and were merged based on student name.

```{r fig-response-time-and-gpa}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of respondents' survey response times and self-reported GPA
#| fig-subcap: ["STA302 distribution of survey response times","STA302 \"What is your GPA?\"","STA304 distribution of survey response times","STA304 \"What is your GPA?\""]
#| layout-ncol: 2

all_data |>
  # summarize(mean = mean(mins_complete/60),
  # standard_dev = sd(mins_complete/60))
  ggplot(aes(x = mins_complete / 60)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Time to complete survey (minutes)", y = "Frequency") +
  theme_minimal()

all_data |>
  # summarize(mean = mean(what_is_your_gpa, na.rm = TRUE),
  #           standard_dev = sd(what_is_your_gpa, na.rm = TRUE))
  ggplot(aes(x = what_is_your_gpa)) +
  geom_histogram(binwidth = 0.1) +
  labs(x = "Self-reported GPA", y = "Frequency") +
  theme_minimal()

sta304_data |>
  # summarize(mean = mean(mins_complete/60),
  # standard_dev = sd(mins_complete/60))
  ggplot(aes(x = mins_complete / 60)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Time to complete survey (minutes)", y = "Frequency") +
  theme_minimal()

sta304_data |>
  # summarize(mean = mean(what_is_your_gpa, na.rm = TRUE),
  #           standard_dev = sd(what_is_your_gpa, na.rm = TRUE))
  ggplot(aes(x = what_is_your_gpa)) +
  geom_histogram(binwidth = 0.1) +
  labs(x = "Self-reported GPA", y = "Frequency") +
  theme_minimal()
```

In STA304 three students took longer than one hour to complete the survey and were removed. After those respondents were removed the average time to complete the survey was 10 minutes and the standard deviation was 8 minutes (@fig-response-time-and-gpa-2). This left 131 respondents that were of use and were merged based on name.

In STA302 there was a wide distribution of self-reported GPAs (@fig-response-time-and-gpa-2). Most responses clustered around a B (3.0 out of 4.0), and the average was 3.06 with a standard deviation of 0.55. One factor that may have affected the range is that the course is required for programs in the Statistics, Mathematics and Computer Science Departments. While there was no reason for the respondents to not report the truth, self-reported GPAs also introduce the possibility of reporting bias. For instance, respondents may have provided their cumulative GPA, their most recent term's GPA, or could have misreported it entirely. The mean and standard deviation of self-reported GPA was the same in STA304 (@fig-response-time-and-gpa-4).

Students from a range of years took both courses, but most respondents were in their 3rd or 4th year of study (@tbl-year-dist). The prerequisite two-course sequence is typically completed by students in their second year, would make it difficult to take either course earlier than their 3rd year.

```{r tbl-year-dist}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: \"What year are you?\"
#| tbl-subcap: ["STA302 responses (Winter 2024)","STA304 responses (Fall 2024)"]

all_data |>
  count(what_year_are_you) |>
  pivot_wider(
    names_from = what_year_are_you,
    values_from = n,
    values_fill = list(n = 0)
  ) |>
  kable()

sta304_data |>
  count(what_year_are_you) |>
  pivot_wider(
    names_from = what_year_are_you,
    values_from = n,
    values_fill = list(n = 0)
  ) |>
  kable()
```

Respondents from STA302 had a varied self-perception of their coding and writing abilities (@fig-selfperception). Most respondents believed it was important to be good at writing, but many were either indifferent or do not like to write (@fig-selfperception-1). Respondents also do not find writing to be particularly easy, which could be associated with the reported relative antipathy toward writing. Most respondents were at least somewhat confident in their own writing abilities, but a substantial number felt otherwise. Although few respondents felt that they were confident in their writing ability, more felt that they were able to catch their mistakes, which could indicate a disconnect between how respondents perceive their work and how the work was evaluated.

```{r fig-selfperception}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Self-perception of coding and writing abilities in STA302 (Winter 2024)
#| fig-subcap: ["\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\"", "\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\""]

@@

Respondent self-perceptions regarding coding proficiency and importance varied (@fig-selfperception-2). There was strong consensus on the perceived importance of coding skills. However, respondents' self-assessed coding abilities and enjoyment were more heterogeneous, with a substantial proportion reporting moderate rather than high levels of ease and enjoyment in coding tasks. Overall, there was a moderate level of confidence among respondents in their overall coding ability, willingness to share code, and capacity to identify errors. Notably, respondents express slightly higher confidence in detecting their own coding mistakes compared to general coding ability or code sharing. These patterns suggest that while respondents have developed some coding self-efficacy, there is still considerable potential for enhancing their perceived competence and comfort across various coding-related activities.

Similar results were also found in STA304 (@fig-sta304-selfperception).

```{r fig-sta304-selfperception}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Self-perception of coding and writing abilities in STA304 (Fall 2024)
#| fig-subcap: ["\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\"", "\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\""]

# Writing
writing_perceptions_long <-
  sta304_data |>
  select(
    writing_is_easy_for_me,
    i_like_to_write,
    i_believe_it_is_important_to_be_a_good_writer,
    when_i_edit_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_writing,
    i_am_confident_in_my_overall_writing_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "writing_is_easy_for_me" ~ "\"Writing is easy for me.\"",
      statements == "i_like_to_write" ~ "\"I like to write.\"",
      statements == "i_believe_it_is_important_to_be_a_good_writer" ~ "\"I believe it is important\nto be a good writer.\"",
      statements == "when_i_edit_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I edit it is easy for\nme to catch my mistakes.\"",
      statements == "i_feel_confident_sharing_my_writing" ~ "\"I feel confident\nsharing my writing.\"",
      statements == "i_am_confident_in_my_overall_writing_ability" ~ "\"I am confident in my\noverall writing ability.\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different\nfrom me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

# Coding
coding_perceptions_long <-
  sta304_data |>
  select(
    coding_is_easy_for_me,
    i_like_to_code,
    i_believe_it_is_important_to_be_a_good_coder,
    when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_code,
    i_am_confident_in_my_overall_coding_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "coding_is_easy_for_me" ~ "\"Coding is easy for me\"",
      statements == "i_like_to_code" ~ "\"I like to code\"",
      statements == "i_believe_it_is_important_to_be_a_good_coder" ~ "\"I believe it is important\n to be a good coder\"",
      statements == "when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I check my code, it is\n easy for me to catch my\nmistakes\"",
      statements == "i_feel_confident_sharing_my_code" ~ "\"I feel confident\n sharing my code\"",
      statements == "i_am_confident_in_my_overall_coding_ability" ~ "\"I am confident in\n my overall coding ability\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different\nfrom me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

Though STA302 respondents' self-perceptions around writing and coding were varied, there was strong consensus (79 per cent) that the use of generative AI tools was appropriate within an academic setting (@tbl-ai-familiarity-usage-1). Most respondents who selected "It depends" (12 per cent) generally found in comments that artificial intelligence tools to be appropriate, though with certain guidelines and rules governing their use. However, in STA304, almost all students (87 per cent) felt that it was ethical and appropriate to use generative AI for school and only 4 per cent felt that it was inappropriate. STA302 ran from January to April 2024, while STA304 ran from September to December 2024.

```{r tbl-ai-familiarity-usage}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: "'To what extent do you think using generative AI tools such as ChatGPT by OpenAI (or equivalents) is ethical and appropriate for schoolwork?'"
#| tbl-subcap: ["Responses from STA302 (Winter 2024)", "Responses from STA304 (Fall 2024)"]

all_data |>
  count(ai_schoolwork_appropriate) |>
  mutate(percentage = n / sum(n) * 100,
         percentage = round(percentage, 0)) |> 
  kable(col.names = c("Ethical & Appropriate for school?", "Number", "Percentage"))

sta304_data |>
  count(ai_schoolwork_appropriate) |>
  mutate(percentage = n / sum(n) * 100,
         percentage = round(percentage, 0)) |> 
  kable(col.names = c("Ethical & Appropriate for school?", "Number", "Percentage"))
```

<!-- To understand the role of LLMs in learning, respondents identified their usage in a more granular way by selecting various pre-defined use cases in the survey (@fig-AI-use-and-usefulness-1). Technical questions and explaining concepts were the two top use cases among respondents. More than half of respondents also used LLMs for quick questions, general knowledge, writing paper code, and checking solutions. Just less than half reported that they used it to write paper content, which could suggest that respondents do not feel confident using it to improve writing. -->

Respondents were also asked to rate the helpfulness of LLMs on various tasks assigned during the course on a 4-point scale of "Did not use" to "Very Helpful" (@fig-AI-use-and-usefulness). To simplify the presentation, responses were grouped into two main categories: "Less Helpful" and "More Helpful." The "Less Helpful" category combines responses where students found the AI either "Not helpful" or did not use it for the task, while the "More Helpful" category includes responses where the LLM was considered "Somewhat helpful" or "Very helpful".
